---
title: "World Bank Data Download and Processing"
author: "Kenith Chan"
date: '2025-04-09'
---

## Introduction

This document outlines the process for downloading and formatting World Bank data using the WDI package in R. The World Development Indicators (WDI) is the World Bank's premier compilation of cross-country data on development, containing over 1,600 indicators for approximately 217 economies.

## Required Packages

First, let's install and load the necessary packages:

```{r packages}
# Install packages if not already installed
if (!require("WDI")) install.packages("WDI")
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("countrycode")) install.packages("countrycode")

# Load packages
library(WDI)
library(tidyverse)
library(countrycode)
```

## Exploring Available Indicators

The WDI package provides a function to search for available indicators:

```{r explore_indicators}
# Search for GDP-related indicators
gdp_indicators <- WDIsearch("gdp")
head(gdp_indicators, 10)

# Search for population-related indicators
pop_indicators <- WDIsearch("population")
head(pop_indicators, 10)
```

## Downloading Key World Bank Indicators

Let's download several important development indicators:

```{r download_indicators}
# Define the indicators we want to download
indicators <- c(
  "NY.GDP.PCAP.PP.KD",   # GDP per capita, PPP (constant 2017 international $)
  "NY.GDP.MKTP.KD.ZG",   # GDP growth (annual %)
  "SP.POP.TOTL",         # Total population
  "SP.DYN.LE00.IN",      # Life expectancy at birth
  "SI.POV.GINI",         # GINI index
  "SE.ADT.LITR.ZS",      # Adult literacy rate
  "SL.UEM.TOTL.ZS"       # Unemployment rate
)

# Define the time period (last 30 years)
start_year <- 1995
end_year <- 2024

# Download the data
wb_data <- WDI(
  indicator = indicators,
  country = "all",
  start = start_year,
  end = end_year,
  extra = TRUE # Include additional country information
)

# Check the structure of the downloaded data
glimpse(wb_data)
```

## Data Cleaning and Preparation

Now let's clean and prepare the data for analysis:

```{r clean_data}
# Rename the columns to more descriptive names
wb_data_clean <- wb_data %>%
  rename(
    gdp_per_capita = NY.GDP.PCAP.PP.KD,
    gdp_growth = NY.GDP.MKTP.KD.ZG,
    population = SP.POP.TOTL,
    life_expectancy = SP.DYN.LE00.IN,
    gini = SI.POV.GINI,
    literacy_rate = SE.ADT.LITR.ZS,
    unemployment_rate = SL.UEM.TOTL.ZS
  )

# Filter out aggregates and keep only countries
wb_data_clean <- wb_data_clean %>%
  filter(region != "Aggregates")

# Check for missing values in key variables
missing_summary <- wb_data_clean %>%
  summarise(across(
    .cols = c(gdp_per_capita, gdp_growth, population, life_expectancy, 
              gini, literacy_rate, unemployment_rate),
    .fns = ~sum(is.na(.)),
    .names = "missing_{.col}"
  ))

print(missing_summary)
```

## Data Analysis Examples

Let's perform some basic analyses with the downloaded data:

```{r analysis}
# Top 10 countries by GDP per capita in the most recent year
top_gdp_countries <- wb_data_clean %>%
  filter(!is.na(gdp_per_capita)) %>%
  filter(year == max(year, na.rm = TRUE)) %>%
  select(country, gdp_per_capita) %>%
  arrange(desc(gdp_per_capita)) %>%
  head(10)

# Top 10 most populous countries
top_pop_countries <- wb_data_clean %>%
  filter(!is.na(population)) %>%
  filter(year == max(year, na.rm = TRUE)) %>%
  select(country, population) %>%
  arrange(desc(population)) %>%
  head(10)

# Display results
print(top_gdp_countries)
print(top_pop_countries)
```

## Saving the Data

Finally, let's save the cleaned data to a file for future use:

```{r save_data}
# Check and adjust working directory if needed
current_dir <- basename(getwd())

if (current_dir == "world_bank") {
  message("Already in world_bank directory")
} else if (current_dir == "scripts") {
  setwd("..")
  message("Changed working directory to world_bank")
} else if (current_dir == "pwb-database") {
  setwd("world_bank")
  message("Changed working directory to world_bank")
} else {
  stop("Please run this script from within the world_bank directory or its scripts subdirectory.")
}

if (!dir.exists("data/processed")) {
  dir.create("data")
  dir.create("data/processed")
}

# Save the cleaned data
saveRDS(wb_data_clean, "data/processed/world_bank_indicators.rds")
write_csv(wb_data_clean, "data/processed/world_bank_indicators.csv")
```

## Conclusion

This document demonstrates how to download, clean, and analyze World Bank data using the WDI package in R. The processed data is saved in CSV and RDS formats for further analysis. The indicators selected provide a comprehensive overview of economic, social, and environmental development across countries and over time.

## Session Info

```{r session_info}
sessionInfo()
```