---
title: "Gallup World Poll Attributes Extraction"
author: "Kenith Chan"
date: "2025-04-02"
output: html_document
---

# Introduction

This document extracts and processes metadata attributes from the Gallup World Poll dataset. The Gallup World Poll contains survey data from respondents across multiple countries and years, covering various well-being and other socioeconomic measures. 

The goal of this script is to:
1. Extract all variable attributes (labels, response options, etc.) from the dataset
2. Identify non-substantive responses (Don't Know, Refused, Not Applicable)
3. Identify binary response questions
4. Combine variable metadata with full question text
5. Save the organized metadata for future use

This metadata will facilitate easier analysis and interpretation of the Gallup World Poll variables.

## Setup

First, we load the necessary libraries for data manipulation.

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
```

## Data Loading

We load the preprocessed Gallup World Poll dataset from an RDS file. The RDS file is converted from the original SAV file because it preserves attributes (like labels and value definitions), is smaller in size, and loads faster in R compared to the SAV format.

```{r}
# Load the preprocessed Gallup World Poll dataset from RDS format
gallupRDS <- readRDS("gallup_world_poll/data/processed/Gallup_World_Poll_022125.rds")
```

## Attribute Extraction

We extract all attributes from each variable in the dataset. These attributes contain valuable metadata such as variable labels and response options. The Gallup World Poll data uses labelled variables to preserve the mapping between numeric values (stored in the data) and their textual meanings (stored as attributes).

This step creates a comprehensive catalog of all variables and their metadata, which will serve as the foundation for our attribute database.

```{r}
# First pass: collect all unique attribute names across all variables
all_attr_names <- c()
for (col_name in names(gallupRDS)) {
  col_attrs <- attributes(gallupRDS[[col_name]])
  all_attr_names <- unique(c(all_attr_names, names(col_attrs)))
}

# Remove standard attributes we're already handling separately (label, labels, class)
extra_attr_names <- setdiff(all_attr_names, c("label", "labels", "class"))

# Create an empty tibble with initial columns for variable name, label, and response options
column_specs <- list(
  QTAG = character(),        # Variable name/identifier
  ShortText = character(),   # Variable label (short description)
  Responses = list()         # List to store value labels
)

# Add columns for any other attributes we found in the dataset
for (attr_name in extra_attr_names) {
  column_specs[[attr_name]] <- list()
}

# Create the tibble with all required columns
gallup_raw_attr <- tibble::tibble(!!!column_specs)

# Loop through all columns in the dataset to extract their attributes
for (col_name in names(gallupRDS)) {
  # Get the column
  col <- gallupRDS[[col_name]]
  
  # Extract all attributes from this column
  col_attrs <- attributes(col)
  
  # Create a list to hold all values for this variable
  row_values <- list(
    QTAG = col_name,
    ShortText = NA_character_,  # Default to NA if no label exists
    Responses = list(NULL)      # Default to NULL if no value labels exist
  )
  
  # Add the variable label if it exists
  if (!is.null(col_attrs$label)) {
    row_values$ShortText <- col_attrs$label
  }
  
  # Add value labels mapping if it exists
  if (!is.null(col_attrs$labels)) {
    # Create a named list with values as keys and labels as values
    responses_list <- as.list(col_attrs$labels)
    names(responses_list) <- names(col_attrs$labels)
    
    # Store the named list directly - it's more natural to work with
    row_values$Responses <- list(responses_list)
  }
  
  # Add any additional attributes to their respective columns
  for (attr_name in extra_attr_names) {
    if (!is.null(col_attrs[[attr_name]])) {
      row_values[[attr_name]] <- list(col_attrs[[attr_name]])
    } else {
      row_values[[attr_name]] <- list(NULL)
    }
  }
  
  # Add this variable's information as a new row in our tibble
  gallup_raw_attr <- gallup_raw_attr %>%
    add_row(!!!row_values)
}
```

## Non-substantive responses (DK, Refused, NA)

Survey data often contains special codes for non-substantive responses - answers that don't provide actual information about the question being asked. These typically include:

1. **Don't Know (DK)**: When respondents are unable to answer
2. **Refused**: When respondents decline to answer
3. **Not Applicable (NA)**: When the question doesn't apply to the respondent

Identifying these responses is critical because they need special handling during analysis. They are not missing at random but represent specific types of non-response that might have substantive meaning.

We use regular expressions to identify these response types across variables, allowing for variations in how they are labeled in the dataset.

```{r}
# Create regex patterns to match different types of non-substantive responses
dk.text <- "(DK\\b|DK\\W)|((Do\\W?not)|dont|don't)\\W?know\\W?$" # Matches various forms of "Don't Know"
refused.text <- "(refuse|refused|RF\\b|RF\\W)"  # Matches various forms of "Refused"
na.text <- "(\\(does\\W?not\\W?apply\\))|(Not\\W?Applicable)|(\\bNA\\b)|(\\(N\\W?A\\))|(No\\W?answer)|(No\\W?response)|(Not\\W?Available\\W?$)" # Matches various forms of NA responses

# Identify and extract non-substantive responses for each variable
gallup_nonsub_attr <- gallup_raw_attr %>%
    rowwise() %>%
    mutate(
        # For "Don't Know" responses: extract both the text label and the corresponding numeric value
        DK_text = ifelse(any(grepl(dk.text, names(Responses), ignore.case = TRUE)),
                         names(Responses)[grep(dk.text, names(Responses), ignore.case = TRUE)], 
                         NA),
        DK_value = ifelse(any(grepl(dk.text, names(Responses), ignore.case = TRUE)), 
                          Responses[[grep(dk.text, names(Responses), ignore.case = TRUE)]], 
                          NA),
        
        # For "Refused" responses: extract text label and numeric value
        Refused_text = ifelse(any(grepl(refused.text, names(Responses), ignore.case = TRUE)), 
                              names(Responses)[grep(refused.text, names(Responses), ignore.case = TRUE)], 
                              NA),
        Refused_value = ifelse(any(grepl(refused.text, names(Responses), ignore.case = TRUE)), 
                               Responses[[grep(refused.text, names(Responses), ignore.case = TRUE)]], 
                               NA),
        
        # For "Not Applicable" responses: extract text label and numeric value
        NA_text = ifelse(any(grepl(na.text, names(Responses), ignore.case = TRUE)),
                                names(Responses)[grep(na.text, names(Responses), ignore.case = TRUE)], 
                                NA),
        NA_value = ifelse(any(grepl(na.text, names(Responses), ignore.case = TRUE)),
                               Responses[[grep(na.text, names(Responses), ignore.case = TRUE)]], 
                               NA)
    ) %>%
    ungroup()

# Print unique instances of each category to verify our pattern matching
# This helps ensure we've correctly identified non-substantive responses
print(unique(gallup_nonsub_attr$DK_text[!is.na(gallup_nonsub_attr$DK_text)]))
print(unique(gallup_nonsub_attr$Refused_text[!is.na(gallup_nonsub_attr$Refused_text)]))
print(unique(gallup_nonsub_attr$NA_text[!is.na(gallup_nonsub_attr$NA_text)]))

# Check for variables without "Don't Know" responses to verify our pattern matching
# This helps identify if we've missed any patterns or variables that might have unusual coding
gallup_nonsub_attr %>%
  filter(!grepl("REG", QTAG) & is.na(DK_text)) %>%  # Exclude region variables
  distinct(Responses, .keep_all = TRUE) %>%          # Look at unique response sets
  mutate(
    Response_options = sapply(Responses, function(x) paste(names(x), collapse = ", ")),  # Create readable list of responses
    Response_values = sapply(Responses, function(x) paste(x, collapse = ", "))          # Create readable list of values
  ) %>%
  select(QTAG, ShortText, Refused_text, Refused_value, NA_text, NA_value, Response_values, Response_options) %>%
  View()  # Open in a spreadsheet-like viewer for manual examination

# Similarly check for variables without "Refused" responses
gallup_nonsub_attr %>%
  filter(!grepl("REG", QTAG) & is.na(Refused_text)) %>%
  distinct(Responses, .keep_all = TRUE) %>%
  mutate(
    Response_options = sapply(Responses, function(x) paste(names(x), collapse = ", ")),
    Response_values = sapply(Responses, function(x) paste(x, collapse = ", "))
  ) %>%
  select(QTAG, ShortText, DK_text, DK_value, NA_text, NA_value, Response_values, Response_options) %>%
  View()

# Combine all non-substantive responses into a single named list for easier handling
# This creates a consolidated record of all non-substantive responses for each variable
gallup_nonsub_attr <- gallup_nonsub_attr %>%
    rowwise() %>%
    mutate(
        NonSubResponses = list({
            ns_responses <- list()
            # Add each type of non-substantive response to our list if it exists
            if(!is.na(DK_text)) ns_responses[[DK_text]] <- DK_value
            if(!is.na(Refused_text)) ns_responses[[Refused_text]] <- Refused_value
            if(!is.na(NA_text)) ns_responses[[NA_text]] <- NA_value
            ns_responses
        })
    ) %>%
    ungroup() %>%
    # Remove the individual columns now that we've consolidated them
    select(-DK_text, -DK_value, -Refused_text, -Refused_value, -NA_text, -NA_value)
```

## Questions with Binary Responses (1 and 2)

Many survey questions in the Gallup World Poll use a binary response format where "1" typically represents a positive or affirmative response (e.g., "Yes", "Agree") and "2" represents a negative response (e.g., "No", "Disagree").

For analytical purposes, it's often more intuitive to recode these as 1 and 0, where 1 represents the positive response and 0 represents the negative. This facilitates interpretation of means (as proportions) and allows for more intuitive modeling.

In this section, we identify which variables follow this specific binary response pattern to facilitate future recoding.

```{r}
# Identify variables with binary 1/2 response options (after excluding non-substantive responses)
gallup_binary_attr <- gallup_nonsub_attr %>%
    rowwise() %>%
    mutate(
        # Create a subset of responses that excludes non-substantive options
        SubstantiveResponses = list(Responses[!names(Responses) %in% names(NonSubResponses)]),
        
        # Check if the variable has exactly two substantive responses coded as 1 and 2
        # This is the classic binary response pattern in many surveys
        BinaryResponses = length(SubstantiveResponses) == 2 && 
                          all(sort(as.numeric(SubstantiveResponses)) == c(1, 2))
    ) %>%
    ungroup() %>%
    # Remove the temporary column since we've captured what we need in BinaryResponses
    select(-SubstantiveResponses)
```

## Add Question Text as a new column

While the variable labels (stored in the "ShortText" column) provide brief descriptions of variables, the actual question text as presented to respondents provides important context. The full question text can help researchers understand nuances that might affect interpretation of the responses.

We merge our metadata with the full question text from [Gallup Reference Tool](https://wpr.gallup.com/). This creates a more complete reference for each variable.

```{r}
# Import the question text file that contains the full text of each survey question
questions_df <- read.csv("gallup_world_poll/data/raw/Questions.csv", stringsAsFactors = FALSE) %>% 
    select(QTAG = QTag, QuestionText = Question.Text)  # Select and rename relevant columns

# Join the question text with our metadata using QTAG as the key
gallup_attr <- gallup_binary_attr %>%
    left_join(questions_df, by = "QTAG") %>%
    # Select and organize our final columns
    select(QTAG, ShortText, QuestionText, Responses, NonSubResponses, BinaryResponses)
```

## Save the final attribute list to an RDS file and a JSON file

We save our compiled metadata in two formats:
1. An RDS file for easy loading in future R analysis
2. A JSON file for interoperability with other programming languages and systems

This makes the metadata accessible for a variety of future uses, including data dictionaries, codebooks, and automated data processing pipelines.

```{r}
# Save as RDS for use in R
saveRDS(gallup_attr, "gallup_world_poll/data/metadata/Gallup_World_Poll_022125_Attributes.rds")
# Save the final attribute list to a JSON file
jsonlite::write_json(gallup_attr, "gallup_world_poll/data/metadata/Gallup_World_Poll_022125_Attributes.json", pretty = TRUE, auto_unbox = TRUE)
```
