---
title: "Gallup World Poll Cleaning Script Generator"
author: "Kenith Chan"
date: "2025-04-02"
output: html_document
---

## Gallup World Poll Cleaning Script Generator

This R Markdown file generates customized data cleaning scripts for Gallup World Poll data based on user specifications.

## Description

This generator creates R scripts for cleaning Gallup World Poll data by:
- Extracting specified variables from raw datasets
- Converting non-substantive responses to NA values
- Recoding binary variables (1=Yes, 2=No) to (1=Yes, 0=No)
- Calculating affect indices (optional)
- Saving the cleaned dataset with appropriate naming

This uses metadata previously extracted using `0_GallupWorldPoll_attributes_extraction.rmd`.

## Template Files

Located in the templates directory:
- `cleaning_script_template.R`: Main template
- `na_conversion_template.R`: NA value conversion
- `binary_conversion_template.R`: Binary response conversion
- `affects_calculation_template.R`: Affects calculation

## Required Folder Structure

```
gallup_world_poll/
  ├── data/
  │   ├── metadata/        # Contains variable metadata
  │   │   └── Gallup_World_Poll_XXXXXX_Attributes.rds
  └── scripts/
      ├── 1_GallupWorldPoll_cleaningScript_generation.rmd  # This script
      ├── generated_scripts/  # Will contain generated cleaning scripts
      └── templates/          # Contains template files for script generation
          ├── affects_calculation_template.R
          ├── binary_conversion_template.R
          ├── cleaning_script_template.R
          └── na_conversion_template.R
```


## Using Generated Scripts

1. **Copy** the generated script to your project
2. **Place** your raw Gallup data file in your data directory
3. **Run** the cleaning script in your environment

## Troubleshooting

- **Missing packages**: Install required packages
- **Template files missing**: Check templates directory
- **Metadata file missing**: Verify metadata path
- **Project name**: Set a project name in parameters

## Quick Start Guide

**Only modify the "User Parameters" section below and run the entire script.**

1. **Install Required Packages** (if needed):
   ```r
   install.packages(c("dplyr", "glue", "rio"))
   ```

2. **Set Parameters** in the "User Parameters" section:
   - `projectName`: Your project identifier (required)
   - `author`: Your name
   - `rawFileName`: Name of your Gallup data file
   - `rawReleaseDate`: Release date
   - `variables`: Variables to include
   - `needAffects`: TRUE for affect calculations

3. **Run All Chunks** (Click "Run All" or press Ctrl+Alt+R)

4. The generated script will be saved to the output directory

```{r load_libraries, message=FALSE, warning=FALSE}
# Check and install required packages if needed
required_packages <- c("dplyr", "glue", "rio")
missing_packages <- required_packages[!required_packages %in% installed.packages()[,"Package"]]

if(length(missing_packages) > 0) {
  stop(paste("Please install the following packages before running this script:", 
             paste(missing_packages, collapse=", "), 
             "\nUse: install.packages(c('", paste(missing_packages, collapse="', '"), "'))", sep=""))
}

# Load necessary libraries
library(dplyr)
library(glue)
library(rio)
```

### User Parameters

```{r user_parameters}
# Parameters to be specified by the user
params <- list(
  projectName = "TEST",  # Project name (will be used in output filenames) - REQUIRED
  author = "Kenith Chan", # Author name for the generated script
  rawFileName = "The_Gallup_022125.dat",  # Name of raw data file
  rawReleaseDate = "March 11, 2025",  # Release date of the raw data file
  metadata_path = "data/metadata/Gallup_World_Poll_022125_Attributes.rds", # Path to metadata file
  needAffects = TRUE,  # Whether to include affect calculations
  variables = c(  # Variables to include in the subset
    "WGT", "COUNTRYNEW", "COUNTRY_ISO3",
    "YEAR_CALENDAR", "WP4", "YEAR_WAVE",
    "EMP_2010", "INCOME_5", "WP1219", "WP1220",
    "WP1223", "WP3117", "WP16", "WP18",
    "WP63", "WP67", "WP69", "WP70", "WP71", "WP74"
  ),

  # You don't need to change the following parameters
  output_dir = "scripts/generated_scripts", # Output directory for generated script
  templates_dir = "scripts/templates", # Directory containing template files
  todayDate = Sys.Date(),  # Current date
  raw_data_dir = "data/raw" # Directory containing raw data files in your repository
)

# Validate parameters
if(params$projectName == "") {
  stop("Please provide a project name in the params list.")
}

# Display the parameters for confirmation
print("Script will be generated with the following parameters:")
print(params)
```


# Important: You typically don't need to modify anything below this line
# ----------------------------------------------------------------


### Load Template Files
```{r load_templates, error=TRUE}
# Check and adjust working directory if needed
current_dir <- basename(getwd())

if (current_dir == "gallup_world_poll") {
  message("Already in gallup_world_poll directory")
} else if (current_dir == "scripts") {
  setwd("..")
  message("Changed working directory to gallup_world_poll")
} else if (current_dir == "pwb-database") {
  setwd("gallup_world_poll")
  message("Changed working directory to gallup_world_poll")
} else {
  stop("Please run this script from within the gallup_world_poll directory or its scripts subdirectory.")
}

# Function to read template files with error handling
read_template <- function(template_path) {
  if (!file.exists(template_path)) {
    stop(paste("Template file not found:", template_path, 
               "\nPlease ensure all template files exist in:", params$templates_dir))
  }
  readLines(template_path, warn = FALSE) |> 
    paste(collapse = "\n")
}

# Define template paths
main_template_path <- file.path(params$templates_dir, "cleaning_script_template.R")
na_conversion_template_path <- file.path(params$templates_dir, "na_conversion_template.R")
binary_conversion_template_path <- file.path(params$templates_dir, "binary_conversion_template.R")
affects_template_path <- file.path(params$templates_dir, "affects_calculation_template.R")

# Try to read templates with detailed error messages
tryCatch({
  main_template <- read_template(main_template_path)
  na_conversion_template <- read_template(na_conversion_template_path)
  binary_conversion_template <- read_template(binary_conversion_template_path)
  affects_template <- read_template(affects_template_path)
  
  message("All template files loaded successfully.")
}, error = function(e) {
  stop(paste("Error loading templates:", e$message,
             "\nPlease check that all required template files exist."))
})
```

### Process Metadata for NA Values and Binary Response Identification

```{r process_metadata, error=TRUE}
# Load the Gallup reference data with better error handling
tryCatch({
  if(file.exists(params$metadata_path)) {
    gallupRef <- readRDS(params$metadata_path)
    message("Metadata loaded successfully.")
  } else {
    stop(paste("Metadata file not found at:", params$metadata_path, 
               "\nPlease run the attributes extraction script first or check the path."))
  }
  
  # Function to generate code for converting specific values to NA
  toNACode <- function(var){
    if(!var %in% gallupRef$QTAG) {
      warning(paste("Variable", var, "not found in metadata. NA conversion will be skipped for this variable."))
      return(NULL)
    }
    
    # Extract the NA values for the variable from NonSubResponses
    var_row <- gallupRef[gallupRef$QTAG == var, ]
    nonSubResponses <- var_row$NonSubResponses[[1]]
    
    if(length(nonSubResponses) > 0) {
      na_values <- unlist(nonSubResponses)
      valuesToNA <- paste(na.omit(na_values), collapse = ", ")
      
      if(nchar(valuesToNA) > 0){
        return(glue("{var} = ifelse({var} %in% c({valuesToNA}), NA, {var})"))
      }
    }
    return(NULL)
  }
  
  # Generate NA conversion code for all variables
  toNACode_list <- lapply(params$variables, toNACode)
  toNACode_filtered <- toNACode_list[!sapply(toNACode_list, is.null)]
  na_conversion_list <- paste(toNACode_filtered, collapse = ",\n    ")
  
  # Check if binary response variables exist and generate code
  # Use the BinaryResponses column from the actual RDS structure
  binaryVars <- gallupRef$QTAG[gallupRef$BinaryResponses == TRUE]
  binaryVars <- intersect(binaryVars, params$variables)
  
  binary_conversion_list <- ""
  if(length(binaryVars) > 0){
    binaryQCode <- function(var){
      return(glue("{var} = ifelse({var} == 2, 0, {var})"))
    }
    binary_conversion_list <- lapply(binaryVars, binaryQCode) %>% 
      paste(collapse = ",\n    ")
  }
  
}, error = function(e) {
  stop(paste("Error processing metadata:", e$message))
})
```

### Generate Complete Cleaning Script

```{r generate_complete_script, error=TRUE}
tryCatch({
  # Format the variables list for the template with ShortText as comments
  variables_list <- sapply(params$variables, function(var) {
    # Find ShortText for the variable
    short_text <- ""
    if(var %in% gallupRef$QTAG) {
      short_text <- gallupRef$ShortText[gallupRef$QTAG == var]
      if(!is.na(short_text) && nchar(short_text) > 0) {
        return(sprintf('"%s" # %s', var, short_text))
      }
    }
    return(sprintf('"%s"', var)) # Return without comment if ShortText not found
  }) %>% paste(collapse = ',\n  ')
  
  # Format date for filenames
  date_format <- format(params$todayDate, "%y%m%d")
  
  # Apply NA conversion template if needed
  na_conversion_code <- ""
  if(length(toNACode_filtered) > 0) {
    na_conversion_code <- glue(na_conversion_template, .open = "{{", .close = "}}", 
                             NA_CONVERSION_LIST = na_conversion_list)
  }
  
  # Apply binary conversion template if needed
  binary_conversion_code <- ""
  if(length(binaryVars) > 0) {
    binary_conversion_code <- glue(binary_conversion_template, .open = "{{", .close = "}}", 
                                 BINARY_CONVERSION_LIST = binary_conversion_list)
  }
  
  # Apply affects calculation template if needed
  affects_calculation <- ""
  if(params$needAffects) {
    affects_calculation <- affects_template
  }
  
  # Apply all replacements to the main template
  cleaningCodeContent <- glue(main_template, 
                            .open = "{{", 
                            .close = "}}",
                            AUTHOR = params$author,
                            TODAY_DATE = params$todayDate,
                            RAW_FILENAME = params$rawFileName,
                            RAW_RELEASE_DATE = params$rawReleaseDate,
                            RAW_DATA_DIR = params$raw_data_dir,
                            VARIABLES_LIST = variables_list,
                            PROJECT_NAME = params$projectName,
                            DATE_FORMAT = date_format,
                            NA_CONVERSION_CODE = na_conversion_code,
                            BINARY_CONVERSION_CODE = binary_conversion_code,
                            AFFECTS_CALCULATION = affects_calculation,
                            SCRIPT_NAME = glue("GWP_cleaningCode_{params$projectName}_{date_format}.Rmd")
                            )
  
  # Display the generated script
  cat("Generated Script Preview (truncated):\n\n")
  cat(substr(cleaningCodeContent, 1, 1000), "...\n")

}, error = function(e) {
  stop(paste("Error generating script:", e$message))
})
```

### Save Generated Script

```{r save_script, error=TRUE}
tryCatch({
  # Create output directory if it doesn't exist
  if (!dir.exists(params$output_dir)) {
    dir.create(params$output_dir, recursive = TRUE)
    message(paste("Created output directory:", params$output_dir))
  }
  
  # Generate filename
  output_filename <- glue("GWP_cleaningCode_{params$projectName}_{date_format}.Rmd")
  output_path <- file.path(params$output_dir, output_filename)
  
  # Write the script to file
  writeLines(cleaningCodeContent, output_path)
  
  # Confirmation message
  message(glue("✓ SUCCESS: Script saved to: {output_path}"))
  
}, error = function(e) {
  stop(paste("Error saving script:", e$message))
})
```

